Escaping RGBland: Selecting Colors for Statistical Graphics

### 2008-10-20 pkg/paper/Palettes/CSDA.Rnw
added refs to CSDA paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@43 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\Abstract{
  Statistical graphics are often augmented by the use of color
  coding information contained in some variable. When this involves
  the shading of areas (and not only points or lines)---e.g., as in bar plots, pie charts,
  mosaic displays or heatmaps---it is important that the colors are perceptually based
  and do not introduce optical illusions or systematic bias. It is discussed
  how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes).
}

@@

A color model typically supported by software packages involves the specification
of colors as Red-Green-Blue (RGB) triplets. However, this specification
corresponds to color generation on a computer screen \citep[see][]{color:Poynton:2000}
rather than corresponding to human color perception. For humans, it is
virtually impossible to control the perceptual properties of a color in this
color space because there is no single dimension that corresponds to, e.g.,
the hue or the brightness of the color. As a consequence, various perceptually-based
color spaces have been suggested, where each dimension of the color space can
be matched with a perceptual property. One approach widely implemented in
software packages involves Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
a simple transformation of RGB triplets \citep[see][]{color:Wiki+HSV:2007}.
Unfortunately, the dimensions in HSV space map poorly to perceptual
properties and the use of HSV colors encourages the use of 
highly saturated colors. A perceptually based color model that avoids
these problems involves Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
which is a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
This is the color space we advocate in this article.

@@

For choosing color palettes, it is helpful to have a basic idea how
human color vision works. It has been hypothesized that it evolved in three
distinct stages: 1.~perception of \emph{light/dark} contrasts (monochrome only),
2.~\emph{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
3.~\emph{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
\cite{color:Ihaka:2003} and \cite{color:Lumley:2006} for more details and references.
Furthermore, physiological studies have shown that light is captured by three
distinct receptors (so-called cones) in the retina and hence coded in three dimensions
by the human visual system. The perception of this coding, i.e., the subjective
experience of this light, is less well-understood; however, current psychophysical
theories all describe perception in terms of three dimensions \citep{color:Knoblauch:2002}.
Therefore, colors are typically described as locations in 3-dimensional spaces.

@@

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{color:CIE:2004}. \cite{color:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, HCL colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$.
%% HCL space has the shape of a distorted double cone: the admissible combinations of
%% chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
%% The lower panel of Figure~\ref{fig:hsv-hcl} shows
%% vertical sections through this space: each of the resulting hue/chroma planes (given
%% luminance) is now properly balanced towards the same gray (going from black to white
%% with increasing luminance) which becomes obvious when converting the colors to
%% a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. 
Given a certain luminance $L$, all colors resulting from different combinations
of hue $H$ and chroma $C$ have the same level of brightness, i.e.,
are balanced towards the same gray and thus look virtually identical
when converted to a gray scale. However, the admissible combinations of
chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
The reason for this is that some hues lead to light and others to dark colors,
e.g., full chroma yellow is brighter (i.e., has higher luminance) than full chroma blue.

@@

@@

>>>>

\Abstract{
  Statistical graphics are often augmented by the use of color
  coding information contained in some variable. When this involves
  the shading of areas (and not only points or lines)---e.g., as in bar plots, pie charts,
  mosaic displays or heatmaps---it is important that the colors are perceptually based
  and do not introduce optical illusions or systematic bias. This article
discusses
  how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes).
}

@@

A color model typically supported by software packages involves the specification
of colors as Red-Green-Blue (RGB) triplets. However, this specification
corresponds to color generation on a computer screen \citep[see][]{color:Poynton:2000}
rather than corresponding to human color perception. For humans, it is
virtually impossible to control the perceptual properties of a color in this
color space because there is no single dimension that corresponds to, e.g.,
the hue or the brightness of the color. As a consequence, various perceptually-based
color spaces have been suggested, where each dimension of the color space can
be matched with a perceptual property. One approach widely implemented in
software packages involves Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
a simple transformation of RGB triplets \citep[see][]{color:Wiki+HSV:2007}.
Unfortunately, the dimensions in HSV space map poorly to perceptual
properties and the use of HSV colors encourages the use of 
highly saturated colors. A perceptually based color model that mitigates
these problems involves Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
which is a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
This is the color space we advocate in this article.

@@

For choosing color palettes, it is helpful to have a basic idea how
human color vision works. It has been hypothesized that it evolved in three
distinct stages: 1.~perception of \emph{light/dark} contrasts (monochrome only),
2.~\emph{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
3.~\emph{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
\cite{color:Kaiser+Boynton:1996}, 
\cite{color:Knoblauch:2002}, \cite{color:Ihaka:2003}, 
and \cite{color:Lumley:2006} for more details and references.
Furthermore, physiological studies have shown that light is captured by three
distinct receptors (so-called cones) in the retina and hence coded in three dimensions
by the human visual system. The perception of this coding, i.e., the subjective
experience of this light, is less well-understood; however, current psychophysical
theories all describe perception in terms of three dimensions \citep{color:Knoblauch:2002}.
Therefore, colors are typically described as locations in 3-dimensional spaces.

@@

To overcome these drawbacks, various color spaces have been suggested
that more closely map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{color:CIE:2004}. 
These color spaces are still not perfect---such an ideal is never
truly attainable because color perception will always
differ between individuals---but they offer a significant
improvement over HSV, as Figure~\ref{fig:wheel} illustrates.

\cite{color:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, HCL colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$.
%% HCL space has the shape of a distorted double cone: the admissible combinations of
%% chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
%% The lower panel of Figure~\ref{fig:hsv-hcl} shows
%% vertical sections through this space: each of the resulting hue/chroma planes (given
%% luminance) is now properly balanced towards the same gray (going from black to white
%% with increasing luminance) which becomes obvious when converting the colors to
%% a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. 
Given a certain luminance $L$, all colors resulting from different combinations
of hue $H$ and chroma $C$ have the same level of brightness, i.e.,
are balanced towards the same gray and thus look virtually identical
when converted to a gray scale. However, the admissible combinations of
chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
The reason for this is that some hues lead to light and others to dark colors,
e.g., full chroma yellow is brighter (i.e., has higher luminance) than full chroma blue.

@@

This section provides several more demonstrations of how 
HCL-based color palettes can be used to produce statistical
plots that are less ``flashy'' and more perceptually balanced,
so that the viewer can
extract information from the plots more easily and more accurately.

@@

The default color palettes in the \pkg{ggplot2} package 
\citep{color:Wickham:2008} are also based on similar ideas to 
those discussed in this article.
